An Open-Source Implementation of the CMPS Algorithm for Assessing Similarity of Bullets

### 2021-10-21 ju-hofmann/ju-hofmann.Rmd
modified the citation in intro and conclusion

According to the Uniform Crime Reporting Program of FBI (Federal Bureau of Investigation), "More than 73 percent (73.7) of the homicides for which the FBI received weapons data in 2019 involved the use of firearms" [@fbiucr], and firearm examination could play an important role in all these cases. \hh{XXX the new data for 2020 came out on Sep 27 - homicides are up by close to 30%. We might want to mention that.}
<!-- Firearm examination has  played an important role in  convictions [@fbiucr] in the United States criminal justice system. \hh{Do we have a number or a percentage of cases where firearm examination has played a role? or maybe gun violence? XXX} --> <!-- https://ucr.fbi.gov/crime-in-the-u.s/2019/crime-in-the-u.s.-2019/topic-pages/expanded-homicide --> 
An important task of firearm examination is to answer the question of whether two pieces of evidence come from the same source or whether a piece of evidence matches a sample obtained from a specific firearm.
Here, in particular, we are interested to determine whether two bullets were fired from the same gun barrel.
Assessing the similarity between two bullets is based on a comparison of striation marks acquired during the firing process as bullets are propelled through the barrel.
Current state of art sees firearms examiners make an assessment of similarity based on a visual comparison, generally, using a comparison microscope [@afte].
This practice has been criticized for its lack of objectivity and the associated problem of determining valid error rates [@pcast].

A report published by the @nrc states that "[m]uch forensic evidence-including, for example, bite marks and firearm and toolmark identification---is introduced in criminal trials without any meaningful scientific validation, determination of error rates, or reliability testing to explain the limits of the discipline." To overcome those criticisms and concerns, researchers have been making an effort to build databases and develop frameworks and algorithms that bring an objective and quantitative assessment into the field \hh{XXX could you summarize the main contribution of each of these papers in at most a sentence?} [@nistdb, @brundage, @hamby, @Hamby:2019, @song2005, @ChumbleyL_Scott2010VoTM, @aoas, @pmid30444940; @cmps].
Note that this is not an exhaustive list of these efforts.
The notion of bullet signatures is one of the results of these efforts.
Bullet signatures can be extracted from bullet land engraved areas (LEAs), typically one for each land engraved area, and how those bullet signatures are extracted from scans will be discussed in the Background section.
Since bullet signatures can numerically describe the striation marks on the bullet and can be understood by algorithms, they played an important role in bringing objectivity into the field.
In the following sections we will: review the background of bullet signature comparisons, discuss how we followed the idea described by @cmps for the implementation of the CMPS algorithm, propose new metrics based on the CMPS score and a principled evaluation framework for algorithmic results comparison,  and present results of applying our implementation to real data.

@@

```{=tex}
\begin{align}
V = \frac{\sum_k n_k (\bar{y}_{k.} - \bar{y}_{..})^2}{\sum_k \sum_l^{n_k} (y_{kl} - \bar{y}_{k.})^2 }.
\end{align}
```
The numerator of the sum of squares ratio $V$ quantifies the variation between the two groups, while the denominator quantifies the variation within each group.
The sum of squares ratio $V$ can be used as an index for evaluating scans, metrics, and different sets of parameters if the same data set is being used.
Some examples will be presented in the following section.
If we impose the normality and independence assumptions on the random effects $e_{kl}$, the sum of squares ratio $V$ becomes a scaled F-statistic with degrees of freedom of $k-1$ and $n_T - k$ and $F = \frac{n_T - k}{k- 1} V$.
If we want to compare different data sets, stating that a certain setup can achieve better separation on one data set than another, we can scale the sum of squares ratio $V$ and obtain the F-statistic to compare the p-values.

Using the sum of squares ratio $V$ as an evaluation metric, we are able to construct a pipeline that aims to find the optimal parameter value for the CMPS algorithm by maximizing the sum of squares ratio.
In the following section, we will use this ratio to compare the CMPS metrics introduced earlier and investigate effects of different sets of parameters.

@@

\wj{Additionally, we propose new metrics based on the CMPS scores and compare the new metrics with the existing metrics. 
We also introduce a principled evaluation framework of algorithmic results using  a measure based on the sum of squares ratio. 
We showcase the implementation with two datasets in Hamby study (Hamby set 252 and Hamby set 44) and compare the CMPS algorithm using different sets of parameters and different metrics, following the evaluation framework based on the sum of squares ratio. }
<!-- Additionally, we apply the \pkg{cmpsR} package to two datasets in Hamby study (Hamby set 252 and Hamby set 44).  -->
The results obtained are promising:  we were both able to reproduce the results in @cmps qualitatively and achieve a clear separation between the known match (KM) comparisons and known non-match (KNM) comparisons in another bullet study.
\wj{However, the comparisons among different sets of parameters suggest that the optimal choice for parameter settings vary between different datasets. We propose an evaluation framework that allows a comprehensive cross-validation of parameter settings used in the algorithm for future research. 
Comparisons of the R implementation of the CMPS algorithm with the random forest model proposed by @aoas  suggest that adding the CMPS score as an additional feature in the random forest model might add further separation between known match (KM) comparisons and known non-match (KNM) comparisons.}

>>>>

<!-- According to the Uniform Crime Reporting Program of FBI (Federal Bureau of Investigation), "more than 73 percent (73.7) of the homicides for which the FBI received weapons data in 2019 involved the use of firearms" [@fbiucr], and firearm examination could play an important role in all these cases.  -->
<!-- \hh{XXX the new data for 2020 came out on Sep 27 - homicides are up by close to 30%. We might want to mention that.} -->
<!-- https://s3-us-gov-west-1.amazonaws.com/cg-d4b776d0-d898-4153-90c8-8336f86bdfec/CIUS/downloads/2020/expanded-homicide-2020.zip -->
<!-- Firearm examination has  played an important role in  convictions [@fbiucr] in the United States criminal justice system. \hh{Do we have a number or a percentage of cases where firearm examination has played a role? or maybe gun violence? XXX} --> <!-- https://ucr.fbi.gov/crime-in-the-u.s/2019/crime-in-the-u.s.-2019/topic-pages/expanded-homicide --> 
\wj{According to the Uniform Crime Reporting Program of FBI (Federal Bureau of Investigation), "more than 76 percent (76.7) of the homicides for which the FBI received weapons data in 2020 involved the use of firearms", and the number of cases that involved the use of firearms in 2020 increased by close to 30 percent (29.3) compared to that in 2019 [@fbiucr]. Firearm examination could play an important role in all these cases. }
An important task of firearm examination is to answer the question of whether two pieces of evidence come from the same source or whether a piece of evidence matches a sample obtained from a specific firearm.
Here, in particular, we are interested to determine whether two bullets were fired from the same gun barrel.
Assessing the similarity between two bullets is based on a comparison of striation marks acquired during the firing process as bullets are propelled through the barrel.
Current state of art sees firearms examiners make an assessment of similarity based on a visual comparison, generally, using a comparison microscope [@afte].
This practice has been criticized for its lack of objectivity and the associated problem of determining valid error rates [@pcast].

<!-- \hh{XXX could you summarize the main contribution of each of these papers in at most a sentence?} -->
A report published by the @nrc states that "[m]uch forensic evidence-including, for example, bite marks and firearm and toolmark identification---is introduced in criminal trials without any meaningful scientific validation, determination of error rates, or reliability testing to explain the limits of the discipline." To overcome those criticisms and concerns, researchers have been making an effort to build databases and develop frameworks and algorithms that bring an objective and quantitative assessment into the field.
\wj{The database} of @nistdb and data sets of @brundage, @hamby, @Hamby:2019 provide great open data resources for research. Algorithms that can be validated and tested and generate quantitative results are developed for 2D and 3D surface texture [@song2005], tool marks [@ChumbleyL_Scott2010VoTM], and striae on Land Engraved Areas (LEAs) of bullets [@aoas, @pmid30444940, @cmps]. 
Note that this is not an exhaustive list of these efforts.
\wj{The notion of bullet signatures is one of the results of these efforts, and we will use it to explain the CMPS algorithm and showcase} the \pkg{cmpsR} implementation.
Bullet signatures can be extracted from bullet land engraved areas (LEAs), typically one for each land engraved area, and how those bullet signatures are extracted from scans will be discussed in the Background section.
Since bullet signatures can numerically describe the striation marks on the bullet and can be understood by algorithms, they played an important role in bringing objectivity into the field.
In the following sections we will: review the background of bullet signature comparisons, discuss how we followed the idea described by @cmps for the implementation of the CMPS algorithm, propose new metrics based on the CMPS score and a principled evaluation framework for algorithmic results comparison,  and present results of applying our implementation to real data.

@@

```{=tex}
\begin{align}
V = \frac{\sum_k n_k (\bar{y}_{k.} - \bar{y}_{..})^2}{\sum_k \sum_l^{n_k} (y_{kl} - \bar{y}_{k.})^2 }.
\end{align}
```
The numerator of the sum of squares ratio $V$ quantifies the variation between the two groups, while the denominator quantifies the variation within each group.
The sum of squares ratio $V$ can be used as an index for evaluating scans, metrics, and different sets of parameters if the same data set is being used.
Some examples will be presented in the following section.
If we impose the normality and independence assumptions on the random effects $e_{kl}$, the sum of squares ratio $V$ becomes a scaled F-statistic with degrees of freedom of $k-1$ and $n_T - k$ and $F = \frac{n_T - k}{k- 1} V$.
If we want to compare different data sets, stating that a certain setup can achieve better separation on one data set than another, we can scale the sum of squares ratio $V$ and obtain the F-statistic and obtain the corresponding p-value as an index for comparison.

Using the sum of squares ratio $V$ as an evaluation metric, we are able to construct a pipeline that aims to find the optimal parameter values for the CMPS algorithm by maximizing the sum of squares ratio.
\wj{Note that other measures of an algorithm, such as the accuracy and the AUC (Area Under the Curve), are also important and useful. But when algorithms achieve 100 percent accuracy and AUC of value 1, we need another measure like the sum of squares ratio to further distinguish the performance of algorithms.} In the following section, we will use the sum of squares ratio to compare the CMPS metrics introduced earlier and investigate effects of different parameter settings.

@@

\wj{Additionally, we propose new metrics based on the CMPS scores and compare the new metrics with the existing metrics. 
We also introduce a principled evaluation framework of algorithmic results using  a measure based on the sum of squares ratio. 
We showcase the implementation with two datasets in Hamby study (Hamby set 252 and Hamby set 44) and compare the CMPS algorithm using different sets of parameters and different metrics, following the evaluation framework based on the sum of squares ratio. }
<!-- Additionally, we apply the \pkg{cmpsR} package to two datasets in Hamby study (Hamby set 252 and Hamby set 44).  -->
The results obtained are promising:  we were both able to reproduce the results in @cmps qualitatively and achieve a clear separation between the known match (KM) comparisons and known non-match (KNM) comparisons in another bullet study.
\wj{However, the comparisons among different sets of parameters suggest that the optimal choice for parameter settings vary between different datasets.}

\wj{Note that Hamby sets 44 and 252 were taken with different resolution scanning devices. The bullets for Hamby 44 and Hamby 252 are fired from the same ten consecutively rifled P-85 Ruger barrels. The difference in parameter settings is therefore quite surprising. As next steps we need validation studies - trying out the CMPS algorithm on firearms and ammunition combinations that are different to test the limits of the algorithm. The evaluation framework we proposed based on the sum of squares ratio will facilitate such studies and other validation of the algorithm.}

\wj{Comparisons of the R implementation of the CMPS algorithm with the random forest model proposed by @aoas  suggest that adding the CMPS score as an additional feature in the random forest model might add further separation between known match (KM) comparisons and known non-match (KNM) comparisons.}
